Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-11-20 rnni_convexity.tex
update description csort

Let us now introduce a representation of caterpillar trees as lists of taxa, which is a simplification of the cluster representation for caterpillar trees and works as follows.
In the list representing a caterpillar tree its taxa are ordered non-decreasing, according to the ranks of their parents.
Note that the only two elements in such a list that have parents of equal rank are the first two elements $l_1$ and $l_2$.
As we assume to have taxon set $X = \{1, \ldots, n\}$, we assume that they are order in the list such that $l_1 < l_2$, according to their ordering as natural numbers.
This ensures that the list representation of caterpillar trees is unique.

@@

The algorithm $\csort$ (Algorithm~\ref{alg:csort}) is a natural modification of the classical \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi}.
Let $T$ and $R$ be the input trees given by their list representations $L_T$ and $L_R$ as introduced above.
The output path $p$ is built stepwise by adding a new list to the end of $p$.
We call the last element of $p$ at any step $L_{T'}$, representing a tree $T'$ and when we say we update $L_{T'}$, we modify this list and append the new version of $L_{T'}$ to $p$.

As in the Bubble Sort algorithm, we go through the list $L_{T'}$ and swap two elements if they appear in different orders in $L_{T'}$ and $L_R$.
Each swap of elements is an update of $T'$ and extends the path $p$ by one element.
We compare pairs of list elements from the start of the list until its end, and repeat this until the whole list is sorted, just as in classical Bubble Sort.
Our algorithm $\csort$ only differs from Bubble Sort when considering the first three elements of the list.
As we discussed above, both the first and second element of $L_{T'}$ can exchange with its third element.
Therefore, we need to consider these moves separately to all other moves.
Most importantly, we need to make sure that these moves are performed in a way that leads to a shortest caterpillar path.
We can guarantee this if the moves are done as described in the following.
Let $x_1, x_2$ and $x_3$ be the first three elements of $L_{T'}$, in this order.
If $x_1$ ($x_2$) is the last element of those three in $L_R$, we update $L_{T'}$ by swapping $x_1$ ($x_2$) and $x_3$.
By doing this we make sure that there will be no point on $p$ where $x_1$ and $x_2$ swap positions, which would cause an additional $\nni$ move that can be avoided if we perform the moves as described.

Because all moves between caterpillar trees are $\nni$ moves, all pairs of taxa that are not in the same order in $L_T$ and $L_R$ need to swap at some point on any path between $L_T$ and $L_R$.
And since only elements that are direct neighbours in a sequence are allowed to swap, with the only exception being the first three elements, the path computed by $\csort$ is a shortest caterpillar path.

@@

The running time of $\csort$ is the same as for Bubble Sort, that is quadratic in $n$.
Within Section~\ref{section:caterpillar_convex} we will use $\csort$ for some proofs.
Therefore it is important to notice that at the end of each loop $i$ starting in Line~\ref{alg:csort:line:loop} of Algorithm~\ref{alg:csort}, the last $i$ elements of $L_{T'}$ are the same as the ones in $L_R$.
This is a well known property of Bubble Sort that $\csort$ inherits from it.

>>>>

The special topology of caterpillar trees gives a natural preorder of taxa that uniquely represents a caterpillar tree.
That is, for a caterpillar tree $T = [\{x_1, x_2\}, \{x_1, x_2, x_3\}, \ldots, \{x_1, \ldots, x_n\}]$ the taxa are preordered according to the ranks of their parents such that $x_1 = x_2 \leq x_3 \leq \ldots \leq x_n$.
From now on when we say that a taxon is below another one we refer to this preorder.
Furthermore, we can use this preorder to create a list representation of caterpillar trees.
For the tree $T$ above this list is $[x_1, x_2, \ldots, x_n]$.
As we consider the taxon set $X = \{1, \ldots, n\}$, we assume that they are ordered in the list such that $x_1 < x_2$, according to their ordering as natural numbers.
This ensures that the list representation of caterpillar trees is unique.

@@

The algorithm $\csort$ (Algorithm~\ref{alg:csort}) is a natural modification of the classical \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi}.
A path $p$ between two given trees $T$ and $R$ is computed by sorting the list representations of $T$ and $R$ as it is done in Bubble Sort.
The general idea is to update the list representing $T$ by considering it from beginning to end and swap two neighboured taxa if they appear in a different order in $R$ than in the current list.
Note that each swap of taxa results in a new tree that is added to the path $p$.
After passing the whole list the last element, which is the uppermost taxon, is at the same position as in $R$.
By repeating this at most $n-1$ times all elements the resulting list represents $R$.

We use this main principle of Bubble Sort for computing shortest caterpillar paths with $\csort$, but there are special consideration needed for the first three taxa.
The reason for this is that both $\nni$ moves on the lowest internal edge of a caterpillar tree result in caterpillar trees.
For computing a shortest caterpillar path the moves on the first three taxa $x_1, x_2, x_3$ need to be done as described in the following.
If $x_1$ ($x_2$) is the uppermost taxon of those three in $R$, we update the current tree by swapping $x_1$ ($x_2$) and $x_3$.
By doing this we make sure that there will be no point on $p$ where $x_1$ and $x_2$ swap positions, as this is an additional $\nni$ move that can be avoided.

Because all moves between caterpillar trees are $\nni$ moves, all pairs of taxa that are not in the same order in $T$ and $R$ need to swap at some point on any path between these trees.
And since only elements that are direct neighbours in a sequence are allowed to swap, with the only exception being the first three elements, the path computed by $\csort$ is a shortest caterpillar path indeed.

@@

The running time of $\csort$ is the same as for Bubble Sort, that is quadratic in $n$.
Though simple, this algorithm will be crucial for some proofs in Section~\ref{section:caterpillar_convex}.
